Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-07-25 paper.tex
Removed sorted blocks. Updates in algorithms.

We adopt the following notations:
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $\alpha_r$: data read rate from disk (in B/s).
\item $\alpha_w$: data write rate to disk (in B/s).
\item $\beta$: disk access time (in s).
\item $\sigma: \mathbb{R}^{+} \rightarrow \mathbb{R}^{+}$: seek time (in s), function of the seek
  distance (in B).
\item $m$: amount of available RAM (in B).
  \item $m'$: amount of used RAM (in B). $m'\leq m$.
\end{itemize}

The following properties are straightforward applications of these
definitions:
\begin{itemize}
\item number of bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, we assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fill a row of the reconstructed
  image: $\sqrt[3]{n}=\nu$.
\end{itemize}
\todo{Make a figure to illustrate the notations.}

@@

We use a simplistic disk model where a disk is characterized by its
read and write rates ($\alpha_r$ and $\alpha_w$, in B/s) and its
access time $\beta$ (in s). In particular, we assume that seeks
require a constant amount of time, regardless of the position seeked
to. In practice, large variations would be expected, but modeling such
variations would inevitably lead to models specific to the hardware,
file system or operating system, which we intentionally avoid
here. Thus, our goal is to find algorithms that minimize the
\emph{number} of seek operations. Likewise, in modern systems, read
and write times are greatly impacted by caching occurring at several
levels, which we do not model here.

@@

\subsection{Sorted blocks}

Sorting blocks alphanumerically by (x,y,z) as shown on Algorithm 2
reduces the seek time between blocks.

\begin{algorithm}[h]
  \caption{Merging from sorted blocks.}
  \label{algo:sorted-blocks}
  \begin{algorithmic}
    \STATE list = sort blocks according to (x,y,z)
    \FOR{each block in list}
      \STATE read block
      \STATE write block in reconstructed image
    \ENDFOR
  \end{algorithmic}
  \end{algorithm}

\todo{Not sure what we are measuring here. The number of seeks will be identical. The difference in seek time is supposed to be very minimal.}

@@

\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}
    \STATE list = sort slices by increasing x values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE slice = list[index]
        \STATE read slice and append it to buffer
        \STATE index += 1
       \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

@@

Cluster reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 4)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (case 1). For case 3 it always returns
true.  Function \texttt{sizeof} (line 8) returns the actual memory
used by the buffer, including only its allocated segments. It should
be noted that writing the buffer to the reconstructed image (line 11)
might involve some manipulation in the buffer to merge its continuous
segments. \todo{Precise this based on Valerie's final code.}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE list = sort blocks by increasing (z,y,x) values
    \STATE initialize buffer
    \STATE (m',case)=\texttt{switch}(m,n,R,b)
    \STATE old\_block = list[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = list[i]
      \STATE overlap = \texttt{check\_overlap}(block,old\_block,case)
      \IF{\texttt{sizeof}(buffer) \textless m' \textbf{and} overlap=false}
      \STATE read block and insert it in buffer
      \ELSE
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
    \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[h]
  \caption{Buffered merging from blocks using multiple reads.}
  \label{algo:multiple-reads}
  \begin{algorithmic}
  \STATE sorted\_split\_name\_list = sort blocks by increasing (x,y,z) values 
  \STATE start\_index = 0, end\_index=(m-1)
  \WHILE{end\_index \textless sizeof(reconstructed\_image)}
    \STATE initialize data\_buffer
    \FOR{split\_name in sorted\_split\_name\_list}
      \IF{indexes of split\_name is in next\_write\_range}
        \STATE read split;
        \STATE extract all indexes in next write range;
        \STATE write the corresponding data to the data\_buffer
      \ENDIF
    \ENDFOR
    \STATE write data\_buffer in reconstructed\_image 
    \STATE clear data\_buffer
    \STATE start\_index = end\_index + 1
    \STATE end\_index += mem
  \ENDWHILE

@@

In multiple reads, a given chunk will be accessed $k$ times and we have: 
\begin{equation*}
   N_\mathrm{multiple\_reads}= nk + \ceil*{\frac{bR}{m}}
\end{equation*}
$k$ depends on the span of memory offsets covered by a chunk in the
reconstructed image, that is, $\left[(d-1)D^2+dD\right]b$.

Finally:
\begin{equation}
   N_\mathrm{multiple\_reads}= n\left( \ceil*{\frac{\left((d-1)D^2+dD\right)b}{m}}\right) + 2\ceil*{\frac{bR}{m}}
\end{equation}

@@

\subsection{Sorted blocks vs blocks?}

>>>>

We adopt the following notations:
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $m$: amount of available RAM (in B).
\item $m'$: amount of used RAM (in B). $m'\leq m$.
\end{itemize}

The following relations are straightforward applications of these
definitions:
\begin{itemize}
\item number of bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, we assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fill a row of the reconstructed
  image: $\sqrt[3]{n}=\nu$.
\end{itemize}
%\todo{Make a figure to illustrate the notations.}

@@

We use a simplistic disk model where a disk is characterized by its
read and write rates and its access time. In particular, we assume
that seeks require a constant amount of time, regardless of the
position seeked to. In practice, large variations would be expected,
but modeling such variations would inevitably lead to models specific
to the hardware, file system or operating system, which we
intentionally avoid here. Thus, our goal is to find algorithms that
minimize the \emph{number} of seek operations. Likewise, in modern
systems, read and write times are greatly impacted by caching
occurring at several levels, which we do not model here.

@@

@@

\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE list = sort slices by increasing x values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = list[index]
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

Cluster reads are described in
Algorithm~\ref{algo:cluster-reads}. Function \texttt{switch} (line 3)
selects one of the three cases based on the amount of available memory
and the number of blocks. It returns \texttt{m'} and \texttt{case},
the identifier of the selected case. Function \texttt{check\_overlap}
(line 7) determines whether two blocks overlap multiple block slices
(case 2) or multiple block rows (case 1). For case 3 it always returns
true.  Function \texttt{sizeof} (line 8) returns the actual memory
used by its argument, including only its allocated segments in case
the argument is a buffer. It should be noted that writing the buffer
to the reconstructed image (line 11) might involve some manipulation
in the buffer to merge its continuous segments. \todo{Precise this
  based on Valerie's final code.}
\begin{algorithm}[h]
  \caption{Buffered merging of blocks with cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}[1]
    \STATE list = sort blocks by increasing (z,y,x) values
    \STATE initialize buffer
    \STATE (m',case)=\texttt{switch}(m,n,R,b)
    \STATE old\_block = list[0]
    \FOR{i = 0 ; i\textless n ; i+=1}
      \STATE block = list[i]
      \STATE overlap = \texttt{check\_overlap}(block,old\_block,case)
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(block) $\geq$ m' \textbf{or} overlap=true}
      \STATE write buffer in reconstructed image
      \STATE clear buffer
      \STATE overlap = false
      \ENDIF
      \STATE read block and insert it in buffer
      \ENDFOR
  \end{algorithmic}
\end{algorithm}

@@

\begin{algorithm}[h]
  \caption{Buffered merging from blocks using multiple reads.}
  \label{algo:multiple-reads}
  \begin{algorithmic}[1]
  \STATE sorted\_split\_name\_list = sort blocks by increasing (z,y,x) values 
  \STATE start\_index = 0, end\_index=(m-1)
  \WHILE{end\_index \textless sizeof(reconstructed\_image)}
    \STATE initialize data\_buffer
    \FOR{split\_name in sorted\_split\_name\_list}
      \IF{indexes of split\_name is in next\_write\_range}
        \STATE read split;
        \STATE extract all indexes in next write range;
        \STATE write the corresponding data to the data\_buffer
      \ENDIF
    \ENDFOR
    \STATE write data\_buffer in reconstructed\_image 
    \STATE clear data\_buffer
    \STATE start\_index = end\_index + 1
    \STATE end\_index += mem
  \ENDWHILE

@@

@@
